---
title: "NBA 2k Stats"
author: "TJ SCANLAN, xyz.sportstats group"
date: ""
output: pdf_document
---
CODE SETUP
---
####Knit Formatting
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
####Library Commands (TJS)
```{r, message=FALSE}
library(dplyr) #loads in dplyr library
library(tidyr) #loads in tidyr library
library(readr) #loads in readr library
library(readxl) #loads in readxl library
library(ggplot2) #loads in ggplot2 library
```

DATA WRANGLING
---
####NBA Data Set Commands (TJS)
```{r}
nba2k_full <- read.csv("nba_rankings_2014-2020.csv") %>% #creates nba_full dataframe from the csv file
  rename(pctFG=FG., threePM=X3PM, threePA=X3PA, pct3P=X3P., pctFT=FT., "+/-"=X..., OVR_RATING=rankings) %>%  #renames all variables with sp. ch.
  group_by(SEASON) %>%  #groups the data by season
  mutate(SHOOTINGpct=((FGM+threePM)/(FGA+threePA)*100))
#glimpse(nba2k_full) #summarizes the data in nba_full
```

####Identifying a player's position using Omri Goldstein's player_data dataset from Kaggle (TJS)
```{r}
player_positions <- read.csv("player_data.csv") #creates player_positions from the .csv file of same name
```

```{r, message=FALSE, warning=FALSE}
player_positions <- player_positions %>% 
  filter(year_end>2013) %>%  #filters out players that retired before the 2014 season
  rename("PLAYER"=name) %>% #renames name column
  separate(position, c("PRIMARY", "SECONDARY")) #seperates position colummn
```

DATA CLEANING
---
####Joining position data to the full dataset (TJS)
```{r}
nba2k_full <- inner_join(player_positions, nba2k_full, by="PLAYER") #manipulates nba2k_full to include position data

nba2k_full$PRIMARY[nba2k_full$PRIMARY == "C"] <- "Center" #renames C to Center
nba2k_full$PRIMARY[nba2k_full$PRIMARY == "F"] <- "Forwards (PF & SF)" #renames F to Forward
nba2k_full$PRIMARY[nba2k_full$PRIMARY == "G"] <- "Guards (PG & SG)" #renames G to Guard
```


RESEARCH QUESTIONS
---
####5) Of the five positions, which occurs most frequently in the top 10 players in each season? Does it change? Are players that play multiple positions better? I believe there will be one position that will be the most frequent across all seasons measured. Possibly positions that have higher shots made will be higher on the list. (TJS)

######Creating tables showing the top 10 players from each season
```{r}
topovr_S1 <- nba2k_full %>% #creates a new frame from S1
  select(SEASON, PLAYER, GP, PRIMARY, SECONDARY, SHOOTINGpct, OVR_RATING) %>% #selects only these columns
  filter(GP>=41, SEASON == "2014-15") %>%  #filters out players that played less than half the season
  arrange(desc(OVR_RATING)) %>%  #arranges the new data frame to display the top players by overall rating
head(10) #keeps the first 10 observations

topovr_S2 <- nba2k_full %>%  #creates a new frame from S2
  select(SEASON, PLAYER, GP, PRIMARY, SECONDARY, SHOOTINGpct, OVR_RATING) %>% #selects only these columns
  filter(GP>=41, SEASON == "2015-16") %>%  #filters out players that played less than half the season
  arrange(desc(OVR_RATING)) %>% #arranges the new data frame to display the top players by overall rating
head(10) #keeps the first 10 observations

topovr_S3 <- nba2k_full %>%  #creates a new frame from S3
  select(SEASON, PLAYER, GP, PRIMARY, SECONDARY, SHOOTINGpct, OVR_RATING) %>% #selects only these columns
  filter(GP>=41, SEASON == "2016-17") %>%  #filters out players that played less than half the season
  arrange(desc(OVR_RATING)) %>%  #arranges the new data frame to display the top players by overall rating
head(10) #keeps the first 10 observations

topovr_S4 <- nba2k_full %>%  #creates a new frame from S4
  select(SEASON, PLAYER, GP, PRIMARY, SECONDARY, SHOOTINGpct, OVR_RATING) %>% #selects only these columns
  filter(GP>=41, SEASON == "2017-18") %>%  #filters out players that played less than half the season
  arrange(desc(OVR_RATING)) %>%  #arranges the new data frame to display the top players by overall rating
head(10) #keeps first 10 rows

topovr_S5 <- nba2k_full %>%  #creates a new frame from S5
  select(SEASON, PLAYER, GP, PRIMARY, SECONDARY, SHOOTINGpct, OVR_RATING) %>% #selects only these columns
  filter(GP>=41, SEASON == "2018-19") %>%  #filters out players that played less than half the season
  arrange(desc(OVR_RATING)) %>%  #arranges the new data frame to display the top players by overall rating
head(10) #keeps first 10 rows

topovr_S6 <- nba2k_full %>%  #creates a new frame from S6
  select(SEASON, PLAYER, GP, PRIMARY, SECONDARY, SHOOTINGpct, OVR_RATING) %>% #selects only these columns
  filter(GP>=36, SEASON == "2019-20") %>%  #filters out players playing less than half the season (19-20 season shortened due to COVID)
  arrange(desc(OVR_RATING)) %>% #arranges the new data frame to display the top players by overall rating
head(10) #keeps first 10 rows
```

######Graphing these relationships
```{r}
q5a <- ggplot(topovr_S1, aes(x = PRIMARY)) + #creates graph from topovr_S1 with primary on x-axis
  geom_bar(fill = "lightblue3") + #makes it a bar graph with a color fill
  labs(title = "Frequency of Each Position in the Top Overall Players 2014-15",
       subtitle = "Category of Positions Only, (i.e. SF and PF are both F)",
       x = "Primary Position",
       y = NULL) + #adds necessary labels for understanding
  theme_bw() #uses bw theme

q5b <- ggplot(topovr_S2, aes(x = PRIMARY)) +
  geom_bar(fill = "lightgreen") +
  labs(title = "Frequency of Each Position in the Top Overall Players 2015-16",
       subtitle = "Category of Positions Only, (i.e. SF and PF are both F)",
       x = "Primary Position",
       y = NULL) +
  theme_bw()

q5c <- ggplot(topovr_S3, aes(x = PRIMARY)) +
  geom_bar(fill = "orangered") +
  labs(title = "Frequency of Each Position in the Top Overall Players 2016-17",
       subtitle = "Category of Positions Only, (i.e. SF and PF are both F)",
       x = "Primary Position",
       y = NULL) +
  theme_bw()

q5d <- ggplot(topovr_S4, aes(x = PRIMARY)) +
  geom_bar(fill = "pink") +
  labs(title = "Frequency of Each Position in the Top Overall Players 2017-18",
       subtitle = "Category of Positions Only, (i.e. SF and PF are both F)",
       x = "Primary Position",
       y = NULL) +
  theme_bw()

q5e <- ggplot(topovr_S5, aes(x = PRIMARY)) +
  geom_bar(fill = "lightgray", color = "red") +
  labs(title = "Frequency of Each Position in the Top Overall Players 2018-19",
       subtitle = "Category of Positions Only, (i.e. SF and PF are both F)",
       x = "Primary Position",
       y = NULL) +
  theme_bw()

q5f <- ggplot(topovr_S6, aes(x = PRIMARY)) +
  geom_bar(fill = "lightgray", color = "blue") +
  labs(title = "Frequency of Each Position in the Top Overall Players 2019-20",
       subtitle = "Category of Positions Only, (i.e. SF and PF are both F)",
       x = "Primary Position",
       y = NULL) +
  theme_bw()
```
######THOUGHTS AND POSSIBLE REVISIONS
I think the graphs turned out pretty well here. To make it easier to compare season-by-season change, I'll want to either combine these sets and add a facet; or, I could find a way to put all of these seperate graphs on one sheet without faceting. The results seem to predict that forwards are more high-caliber players, at least based on shooting percentage. This trend continues throughout all seasons measured, however, in the 17-18 season, forwards and guards had equal presence in the top 10 players by overall. Maybe, to get a better idea of the distribution, I should measure the top 20 players instead of top 10.

####6) What is the relationship between a player’s point differential, time on-court, and their shooting percentage? Could a relationship between these indicate a generalization of their overall? I’m guessing that there will be a positive correlation between point differential and shooting percentage, but since there are so many statistics going into the overall rating, it will be tough to make any assumptions as to how the overall is affected. (TJS)

```{r, message=FALSE}
pointdiff_toc_sp <- nba2k_full %>% #creates new data frame fromnba2k_full
  select(SEASON, PLAYER, "+/-", MIN, SHOOTINGpct, PRIMARY) %>% #selects only these variables
  rename(pointdiff="+/-") %>% #renames the point differential column
  drop_na() %>%  #drops missing values
  filter(SHOOTINGpct != 100 & SHOOTINGpct != 0) #filters out these for accuracy (players with these stats played very few games due to injury or something else)

q6a <- ggplot(data = pointdiff_toc_sp, aes(x = pointdiff, y = SHOOTINGpct)) + #creates graph of pointdiff_toc_sp with these paramaters
  theme_bw() + #sets a theme
  geom_point(aes(color = PRIMARY), alpha = 0.33) + #makes a scatterplot with 0.08 alpha level
  geom_smooth(color = "black", method = "lm", se = 0) + #adds a red linear regression line w/o standard error envelope
  facet_grid(PRIMARY~SEASON) + #facets the data by season
  labs(title = "Comparing Point Differential and Shooting Percentage",
       subtitle = "Faceted by season (2014-2019) and position with 0% and 100% Shooting Omitted",
       caption = "Sources: kaggle.com, basketball-reference.com",
       x = "Point Differential",
       y = "Shooting Percentage",
       color = "Position") #adds necessary labels

q6b <- ggplot(data = pointdiff_toc_sp, aes(x = pointdiff, y = SHOOTINGpct)) + #creates graph of pointdiff_toc_sp with these paramaters
  theme_bw() + #sets a theme
  geom_point(aes(color = PRIMARY), alpha = 0.33) + #makes a scatterplot with 0.08 alpha level
  geom_smooth(color = "black", method = "lm", se = 0) + #adds a red linear regression line w/o standard error envelope
  facet_grid(~SEASON) + #facets the data by season
  labs(title = "Comparing Point Differential and Shooting Percentage",
       subtitle = "Faceted by season (2014-2019) with 0% and 100% Shooting Omitted",
       caption = "Sources: kaggle.com, basketball-reference.com",
       x = "Point Differential",
       y = "Shooting Percentage",
       color = "Position") #adds necessary labels

q6c <- ggplot(pointdiff_toc_sp, aes(y = MIN, x = pointdiff)) + #creates the time on court and point differential graph
  theme_bw() + #uses the bw theme
  geom_point(aes(color = PRIMARY), alpha = 0.4) + #makes it a scatterplot with color indicating position
  geom_smooth(color = "black", method = "lm", se = 0) + #adds a linear regression model
  facet_grid(~SEASON) + #facets by season and primary position
  labs(title = "Correlation between Playing Time and Point Differential",
       subtitle = "Faceted by Season (2014-2019)",
       caption = "Source: kaggle.com, basketball-reference.com",
       y = "Playing Time (min)",
       x = "Point Differential",
       color = "Position")
```

#```{r}
#exporting graphs as images to copy
ggpubr::ggexport(q5a, filename = "fig5a.png")
ggpubr::ggexport(q5b, filename = "fig5b.png")
ggpubr::ggexport(q5c, filename = "fig5c.png")
ggpubr::ggexport(q5d, filename = "fig5d.png")
ggpubr::ggexport(q5e, filename = "fig5e.png")
ggpubr::ggexport(q5f, filename = "fig5f.png")

ggpubr::ggexport(q6a, filename = "fig6a.png")
ggpubr::ggexport(q6b, filename = "fig6b.png")
ggpubr::ggexport(q6c, filename = "fig6c.png")
#```


######THOUGHTS AND POSSIBLE REVISIONS
I think these graphs also turned out well. They should be easier to look at once they are spread out on a larger screen. I might consider changing the interval on the x-axis to shorten it, but in the 2014-15 season there seems to be an outlier at the left side of the chart. I'll have to look into that. Other than that, I think these are close to ready.
